---
title: "Beadbeating input scale"
author: "TBNJ"
date: "2023-04-19"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo    = F,
  # fig.pos = "H",
  message = FALSE,
  warning = FALSE
)

library(here)
setwd(here("Bead_beating"))

source(here("Bead_beating", "scripts/loaddata_scale.R"))
source(here("Bead_beating", "scripts/functions.R"))
source(here("Bead_beating", "scripts/article_theme.R"))

library(tidyverse)
library(ampvis2)
library(vegan)
library(patchwork)
library(ggpubr)
library(DESeq2)
library(rlist)

colors <- c("grey", "darkgoldenrod2", "royalblue4", "seagreen", "salmon")
colors_scale <- c("#C6DBEF", "#4292C6", "#08306B")
```

## Filter Data

> RPM = 1000 removed

```{r remove blanks and pcr controls}
dat.types.amp <- ampvis.amplicon %>%
  amp_subset_samples(!is.na(project_id) & !soil_type %in% c("Blank", "Activated sludge"))

# Create subsets
dat.types.sub3000.amp <- dat.types.amp %>% amp_subset_samples(., minreads = 3000)

rarefy <- min(colSums(dat.types.sub3000.amp[["abund"]]))

set.seed(123)
dat.types.sub3000.amp.ra <- dat.types.sub3000.amp %>%
  amp_subset_samples(., rarefy = rarefy)

dat.types.sub3000.amp.ra.rel_0.1 <- dat.types.sub3000.amp.ra %>%
  amp_subset_samples(normalise = TRUE) %>%
  filter_otus(0.1)

soil <- unique(dat.types.sub3000.amp$metadata$soil_type) %>% sort(.) %>% as.factor(.)
```

We set a cutoff of at 3000 reads, and rarefy to the lowest number of reads in the remaining dataset. Doing this we go from `r nrow(dat_types.amp$metadata)` to `r nrow(dat.types.sub3000.amp$metadata)` samples and create a rarified dataset as well.

### Bray Curtis variation
# Only OTUs above 0.1 % relative abundance is used

```{r}
amp.rarecurve <- dat.types.sub3000.amp %>%
  amp_rarecurve(color_by = "sample_target",
                facet_by = "soil_type",
                facet_scales = "free_x",
                stepsize = 100) +
  scale_color_manual(values = c(colors_scale[c(1,3)]))

tmp <- t(dat.types.sub3000.amp.ra.rel_0.1$abund)

bray_distance <- as.matrix(vegan::vegdist(vegan::decostand(tmp, method = "hellinger"))) %>%
    as.data.frame() %>%
    rownames_to_column("seq_id")
  
metadata <- dat.types.sub3000.amp$metadata %>%
  select(seq_id, kit, sample_target, soil_type)

metadata_renamed <- metadata %>%
  dplyr::rename(
    comparison = seq_id,
    compared_soil = soil_type,
    compared_sample_target = sample_target)

bray_with_metadata <- metadata %>%
  left_join(bray_distance) %>%
  pivot_longer(!seq_id:soil_type, names_to = "comparison", values_to = "value") %>%
  left_join(metadata_renamed) %>%
  filter(value != 0) %>%
  distinct(value, .keep_all = TRUE)
  
bray_variation_sample_target <- bray_with_metadata %>%
  filter(seq_id != comparison) %>%
  filter(paste0(sample_target, soil_type) == paste0(compared_sample_target, compared_soil)) %>%
  filter(!duplicated(paste0(pmax(seq_id, comparison), pmin(seq_id, comparison)))) %>%
  group_by(soil_type, sample_target) %>%
  summarise(
    mean_bray = format(round(mean(value, na.rm = TRUE), 2), nsmall = 2),
    sd_bray = format(round(sd(value, na.rm = TRUE), 2), nsmall = 2), .groups = "drop")

bray_variation_between_sample_targets <- bray_with_metadata %>%
  filter(seq_id != comparison, sample_target != compared_sample_target) %>%
  filter(soil_type == compared_soil) %>%
  filter(!duplicated(paste0(pmax(seq_id, comparison), pmin(seq_id, comparison)))) %>%
  group_by(soil_type) %>%
  summarise(
    mean_bray = format(round(mean(value, na.rm = TRUE), 2), nsmall = 2),
    sd_bray = format(round(sd(value, na.rm = TRUE), 2), nsmall = 2),
    bray_between = paste0(mean_bray, " (", sd_bray, ")")
  ) %>%
  mutate(sample_target = "50") %>%
  select(soil_type, sample_target, bray_between)
```


### Shannon Diversity variation
```{r}
tab_reads <- data.frame(num_reads = apply(dat.types.sub3000.amp$abund, 2, sum)) %>%
  rownames_to_column(var = "seq_id")

lab_metadata <- metadata.amplicon %>%
  select(seq_id, kit, rpm, sample_target, soil_type, conc_qubit, `260/280`, `260/230`, lib_conc) %>%
  filter(!soil_type %in% c("Blank", "Activated sludge"), rpm != "1000") %>%
  mutate(yield_ng = if_else(kit == "PowerSoil Pro HT", conc_qubit * 120, conc_qubit * 60)) %>%
  select(-rpm, -conc_qubit)

rarefied_alphadiversity <- dat.types.sub3000.amp.ra %>%
  amp_alphadiv(
    measure = c("observed", "shannon", "simpson"),
    richness = T
  ) %>%
  dplyr::rename(shannon = Shannon,
                observed_asv = ObservedOTUs) %>%
  select(seq_id, sample_target, soil_type, observed_asv, shannon) %>%
  left_join(tab_reads)

# Join different data: Lab, Alpha, beta.
combine_alpha_beta_with_metadata <- rarefied_alphadiversity %>%
  # add library_concentration
  right_join(lab_metadata) %>%
  # Calculate failed libraries
  group_by(soil_type, sample_target) %>%
  mutate(libs = ifelse(is.na(num_reads) == TRUE, 3-length(is.na(num_reads)), n())) %>%
  # Calculate mean and sd for all numeric varaibles.
  summarise(across(where(is.numeric), .fns = list(mean = ~ mean(., na.rm = T), 
                                                  sd = ~ sd(., na.rm = T)), .names = "{.fn}_{.col}")) %>%
  # Format mean and sd
  ungroup() %>%
  mutate(libs = as.character(mean_libs)) %>%
  mutate(
    libs = as.character(mean_libs),
    across(mean_num_reads:sd_num_reads, ~ format(round(., 0))),
    across(mean_observed_asv:sd_observed_asv, ~ format(round(., 0))),
    across(mean_lib_conc:sd_lib_conc, ~ format(round(., 2))),
    #across(mean_integrity:sd_integrity, ~ format(round(./1000, 2))),
    mean_yield_µg = format(round(as.numeric(mean_yield_ng)/1000, 2)),
    sd_yield_µg = format(round(as.numeric(sd_yield_ng)/1000, 2))) %>%
  select(-mean_yield_ng, -sd_yield_ng) %>%
  # Add bray variation
  left_join(bray_variation_sample_target) %>%
  arrange(soil_type, sample_target) %>%
  mutate(across(where(is.numeric), ~ format(round(., 2), nsmall = 2))) %>%
  select(-c(sd_libs, mean_libs))

```


## Final table
```{r}
mean_sd_table <- combine_alpha_beta_with_metadata %>%
  pivot_longer(cols = -c(soil_type, sample_target, libs, contains("sd")), 
               names_to = "mean", values_to = "mean_value", names_prefix = "mean_") %>%
  pivot_longer(cols = -c(soil_type, sample_target, libs, contains("mean")), 
               names_to = "sd", values_to = "sd_value", names_prefix = "sd_") %>%
  group_by(soil_type, sample_target) %>%
  # Filter for wrong match
  filter(mean == sd) %>%
  #mutate(across(everything(), ~trimws(.))) %>%
  # Paste mean (sd)
  mutate(mean_sd = paste0(mean_value, " (", sd_value, ")")) %>%
  select(!mean_value:sd_value) %>%
  # make wide and add bray variation between sample_targets
  pivot_wider(names_from = mean, values_from = mean_sd) %>%
  left_join(bray_variation_between_sample_targets) %>%
  # Remove duplicate Soil type
  mutate(bray_between = if_else(is.na(bray_between), "", bray_between)) %>%
  # Relocate columns
  relocate(lib_conc, .after = libs) %>%
  ungroup() %>%
  group_by(soil_type) %>%
  #separate_wider_delim(sample_target, delim = "-", names = c("rpm", "time")) %>%
  #mutate_at(vars(soil_type:sample_target), ~ replace(., duplicated(.), "")) %>%
  relocate(c(libs, lib_conc, num_reads, observed_asv, shannon, bray), .after = `260/230`) %>%
  relocate(yield_µg, .after = sample_target)


names <- tibble(
  old_names = colnames(mean_sd_table),
  new_names = c("Soil type", "Input amount [mg]", "DNA yield [µg]", "260/280", "260/230", "Libraries", "Library conc. [ng/µL]", "Number of reads", "Observed ASVs", "Shannon Diversity", "Bray-Curtis Dissimilarity", "BC Dissimilarity\nBetween Input")
)

final_table <- mean_sd_table %>%
  rename_at(vars(names$old_names), ~ names$new_names)

hjustification_matrix <- as.vector(matrix(c(0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1), ncol = ncol(mean_sd_table), nrow = nrow(mean_sd_table), byrow = TRUE))

xjustification_matrix <- as.vector(matrix(c(0.1, 0.1, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9), ncol = ncol(mean_sd_table), nrow = nrow(mean_sd_table), byrow = TRUE))

alpha_table <- final_table %>%
  ggtexttable(rows = NULL, theme = ttheme(colnames.style = colnames_style(fill = "white"), tbody.style = tbody_style(fill = "white", hjust = hjustification_matrix, x = xjustification_matrix))) %>%
  tab_add_hline(at.row = c(1:2), row.side = "top", linewidth = 2) %>%
  tab_add_hline(at.row = seq(4, 10, 2), row.side = "top", linewidth = 1)

alpha_table

#ggsave(here("Bead_beating", "figures", "table_input_scale.png"), plot = alpha_table, device = "png", height = 12, width = 19)

writexl::write_xlsx(final_table, here("Bead_beating", "figures", "subfile2_Table1.xlsx"))
```


## Statistical tests - ANOVA and Mann-Whitney
```{r}
alpha.all <- rarefied_alphadiversity %>%
  right_join(lab_metadata) %>%
  arrange(shannon) %>%
  mutate(shannon_rank = seq(1:length(shannon))) %>%
  arrange(yield_ng) %>%
  mutate(yield_ng_rank = seq(1:length(yield_ng))) %>%
  arrange(`260/280`) %>%
  mutate(`260/280_rank` = seq(1:length(`260/280`))) %>%
  arrange(`260/230`) %>%
  mutate(`260/230_rank` = seq(1:length(`260/230`))) %>%
  mutate(across(soil_type:sample_target, ~as.factor(.)))
  
comp.bray <- bray_with_metadata %>%
  dplyr::rename(bray = value) %>%
  filter(bray != 0) %>%
  filter(soil_type == compared_soil, sample_target == compared_sample_target) %>%
  arrange(bray) %>%
  mutate(bray_rank = seq(1:length(bray))) %>%
  mutate(across(c(sample_target,soil_type,compared_sample_target, compared_soil), ~as.factor(.)))

# Test for normality
shapiro.test(alpha.all$shannon)
qqnorm(alpha.all$shannon)
qqline(y = alpha.all$shannon)

bartlett.test(shannon ~ sample_target, data = alpha.all)
fligner.test(shannon ~ sample_target, data = alpha.all)
car::leveneTest(shannon ~ sample_target, data = alpha.all)

bartlett.test(shannon ~ soil_type, data = alpha.all)
fligner.test(shannon ~ soil_type, data = alpha.all)
car::leveneTest(shannon ~ soil_type, data = alpha.all)

shapiro.test(alpha.all$yield_ng)
qqnorm(alpha.all$yield_ng)
qqline(y = alpha.all$yield_ng)

bartlett.test(yield_ng ~ sample_target, data = alpha.all)
fligner.test(yield_ng ~ sample_target, data = alpha.all)
car::leveneTest(yield_ng ~ sample_target, data = alpha.all)

bartlett.test(yield_ng ~ soil_type, data = alpha.all)
fligner.test(yield_ng ~ soil_type, data = alpha.all)
car::leveneTest(yield_ng ~ soil_type, data = alpha.all)

shapiro.test(alpha.all$`260/280`)
qqnorm(alpha.all$`260/280`)
qqline(y = alpha.all$`260/280`)

bartlett.test(`260/280` ~ sample_target, data = alpha.all)
fligner.test(`260/280` ~ sample_target, data = alpha.all)
car::leveneTest(`260/280` ~ sample_target, data = alpha.all)

bartlett.test(`260/280` ~ soil_type, data = alpha.all)
fligner.test(`260/280` ~ soil_type, data = alpha.all)
car::leveneTest(`260/280` ~ soil_type, data = alpha.all)

shapiro.test(alpha.all$`260/230`)
qqnorm(alpha.all$`260/230`)
qqline(y = alpha.all$`260/230`)

bartlett.test(`260/230` ~ sample_target, data = alpha.all)
fligner.test(`260/230` ~ sample_target, data = alpha.all)
car::leveneTest(`260/230` ~ sample_target, data = alpha.all)

bartlett.test(`260/230` ~ soil_type, data = alpha.all)
fligner.test(`260/230` ~ soil_type, data = alpha.all)
car::leveneTest(`260/230` ~ soil_type, data = alpha.all)

shapiro.test(comp.bray$bray)
qqnorm(comp.bray$bray)
qqline(y = comp.bray$bray)

bartlett.test(bray ~ sample_target, data = comp.bray)
fligner.test(bray ~ sample_target, data = comp.bray)
car::leveneTest(bray ~ sample_target, data = comp.bray)

bartlett.test(bray ~ soil_type, data = comp.bray)
fligner.test(bray ~ soil_type, data = comp.bray)
car::leveneTest(bray ~ soil_type, data = comp.bray)

# ANOVA yield
aov.yield <- lm(yield_ng ~ sample_target * soil_type, data = alpha.all)

aov.yield %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.yield))$sample_target
TukeyHSD(aov(aov.yield))$soil_type
TukeyHSD(aov(aov.yield))$`sample_target:soil_type`

data.yield <- alpha.all %>% filter(!is.na(yield_ng)) %>% pull(yield_ng)
residuals.yield <- resid(aov.yield)
df.yield <- cbind(data.yield, residuals.yield) %>% as.data.frame(.)

ggplot(aes(x = data.yield,
           y = residuals.yield),
       data = df.yield) +
  geom_point() +
  geom_hline(yintercept = 0)

aov.yield.ranks <- lm(yield_ng_rank ~ sample_target * soil_type, data = alpha.all)

aov.yield.ranks %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.yield.ranks))$sample_target
TukeyHSD(aov(aov.yield.ranks))$soil_type
TukeyHSD(aov(aov.yield.ranks))$`sample_target:soil_type`

data.yield.rank <- alpha.all %>% filter(!is.na(yield_ng_rank)) %>% pull(yield_ng_rank)
residuals.yield.rank <- resid(aov.yield.ranks)
df.yield.rank <- cbind(data.yield.rank, residuals.yield.rank) %>% as.data.frame(.)

ggplot(aes(x = data.yield.rank,
           y = residuals.yield.rank),
       data = df.yield.rank) +
  geom_point() +
  geom_hline(yintercept = 0)

# Non-parametric test for differences in purity
ratio1.50 <- alpha.all %>% filter(sample_target == "50") %>% pull(`260/280`)

ratio1.125 <- alpha.all %>% filter(sample_target == "125") %>% pull(`260/280`)

ratio2.50 <- alpha.all %>% filter(sample_target == "50") %>% pull(`260/230`)

ratio2.125 <- alpha.all %>% filter(sample_target == "125") %>% pull(`260/230`)

test.ratio1 <- wilcox.test(ratio1.50, ratio1.50, alternative = "two.sided")
test.ratio1$p.value
# Add small jitter due to ties
d1 = ratio1.50-ratio1.125
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val

test.ratio2 <- wilcox.test(ratio2.50, ratio2.125, alternative = "two.sided")
test.ratio2$p.value
# Add small jitter due to ties
d2 = ratio2.50-ratio2.125
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val

p.adj_ratio <- p.adjust(c(test.ratio1$p.value, test.ratio2$p.value), method = "bonferroni")
p.adj_ratio

# ANOVA shannon
aov.shannon <- lm(shannon ~ sample_target * soil_type, data = alpha.all)

aov.shannon %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.shannon))$sample_target

data.shannon <- alpha.all %>% filter(!is.na(shannon)) %>% pull(shannon)
residuals.shannon <- resid(aov.shannon)
df.shannon <- cbind(data.shannon, residuals.shannon) %>% as.data.frame(.)

ggplot(aes(x = data.shannon,
           y = residuals.shannon),
       data = df.shannon) +
  geom_point() +
  geom_hline(yintercept = 0)

aov.shannon.ranks <- lm(shannon_rank ~ sample_target * soil_type, data = alpha.all)

aov.shannon.ranks %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.shannon.ranks))$sample_target

data.shannon.rank <- alpha.all %>% filter(!is.na(shannon_rank)) %>% pull(shannon_rank)
residuals.shannon.rank <- resid(aov.shannon.ranks)
df.shannon.rank <- cbind(data.shannon.rank, residuals.shannon.rank) %>% as.data.frame(.)

ggplot(aes(x = data.shannon.rank,
           y = residuals.shannon.rank),
       data = df.shannon.rank) +
  geom_point() +
  geom_hline(yintercept = 0)

# ANOVA bray
aov.bray <- lm(bray ~ sample_target * soil_type, data = comp.bray)

aov.bray %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.bray))$sample_target

data.bray <- comp.bray %>% filter(!is.na(bray)) %>% pull(bray)
residuals.bray <- resid(aov.bray)
df.bray <- cbind(data.bray, residuals.bray) %>% as.data.frame(.)

ggplot(aes(x = data.bray,
           y = residuals.bray),
       data = df.bray) +
  geom_point() +
  geom_hline(yintercept = 0)

aov.bray.ranks <- lm(bray_rank ~ sample_target * soil_type, data = comp.bray)

aov.bray.ranks %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.bray.ranks))$sample_target

data.bray.rank <- comp.bray %>% filter(!is.na(bray_rank)) %>% pull(bray_rank)
residuals.bray.rank <- resid(aov.bray.ranks)
df.bray.rank <- cbind(data.bray.rank, residuals.bray.rank) %>% as.data.frame(.)

ggplot(aes(x = data.bray.rank,
           y = residuals.bray.rank),
       data = df.bray.rank) +
  geom_point() +
  geom_hline(yintercept = 0)

# Non-parametric test for differences in Bray dissimilarity
bray_protocol <- bray_with_metadata %>%
  dplyr::rename(bray = value) %>%
  filter(bray != 0) %>%
  filter(soil_type == compared_soil, !sample_target == compared_sample_target)

bray.BS <- comp.bray %>% filter(soil_type == "Beach Sand") %>% pull(bray)
shapiro.test(bray.BS)
bray.CL <- comp.bray %>% filter(soil_type == "Clay") %>% pull(bray)
shapiro.test(bray.CL)
bray.OR <- comp.bray %>% filter(soil_type == "Organic") %>% pull(bray)
shapiro.test(bray.OR)
bray.SA <- comp.bray %>% filter(soil_type == "Sand") %>% pull(bray)
shapiro.test(bray.SA)
bray.SC <- comp.bray %>% filter(soil_type == "Sand-Clay") %>% pull(bray)
shapiro.test(bray.SC)

bray.protocol.BS <- bray_protocol %>% filter(soil_type == "Beach Sand") %>% pull(bray)
shapiro.test(bray.protocol.BS)
bray.protocol.CL <- bray_protocol %>% filter(soil_type == "Clay") %>% pull(bray)
shapiro.test(bray.protocol.CL)
bray.protocol.OR <- bray_protocol %>% filter(soil_type == "Organic") %>% pull(bray)
shapiro.test(bray.protocol.OR)
bray.protocol.SA <- bray_protocol %>% filter(soil_type == "Sand") %>% pull(bray)
shapiro.test(bray.protocol.SA)
bray.protocol.SC <- bray_protocol %>% filter(soil_type == "Sand-Clay") %>% pull(bray)
shapiro.test(bray.protocol.SC)

test.bray.BS <- wilcox.test(bray.BS, bray.protocol.BS, alternative = "two.sided")
test.bray.CL <- wilcox.test(bray.CL, bray.protocol.CL, alternative = "two.sided")
test.bray.OR <- wilcox.test(bray.OR, bray.protocol.OR, alternative = "two.sided")
test.bray.SA <- wilcox.test(bray.SA, bray.protocol.SA, alternative = "two.sided")
test.bray.SC <- wilcox.test(bray.SC, bray.protocol.SC, alternative = "two.sided")

p.bray <- c(test.bray.BS$p.value, test.bray.CL$p.value, test.bray.OR$p.value, test.bray.SA$p.value, test.bray.SC$p.value)
p.adj_bray <- p.adjust(p.bray, method = "bonferroni")
p.adj_bray
```

## Heatmap
```{r}
# Find all phyla in amplicon data
# heatmap.amplicon_phylum <- dat.types.sub3000.amp %>%
#   amp_heatmap(group_by = "seq_id", 
#               min_abundance = 0.001,
#               tax_show = 1000,
#               tax_aggregate = "Phylum", 
#               normalise = FALSE,
#               showRemainingTaxa = FALSE, 
#               plot_values = TRUE,
#               textmap = TRUE) %>%
#   rownames_to_column(var = "Phylum") %>%
#   mutate(across(Phylum, ~ str_replace_all(., "^.*ASV.*$", "Unclassified"))) %>%
#   group_by(Phylum) %>%
#   summarise(across(everything(), ~ sum(.))) %>%
#   ungroup() %>%
#   column_to_rownames(var = "Phylum")

# phylum.text.amp <- heatmap.amplicon_phylum %>%
#   mutate(total = rowSums(.)) %>%
#   filter(!total == 0) %>%
#   select(total) %>%
#   arrange(desc(total)) %>%
#   row.names(.)

# phylum.text.renamed <- gsub("SAR324_cladeMarine_group_B", "\\*SAR324", phylum.text.amp)

# Load new genus ampvis object
# ampvis.sub_phylum <- amp_load(otutable = heatmap.amplicon_phylum,
#                               metadata = metadata.amplicon)

# plot.heatmap_phylum <- ampvis.sub_phylum %>%
#   amp_heatmap(group_by = "sample_target", 
#               min_abundance = 0.1,
#               facet_by = c("soil_type"),
#               tax_aggregate = "OTU",
#               normalise = TRUE,
#               tax_show = phylum.text.amp[c(1:12, 14:26)], 
#               order_y_by = rev(phylum.text.amp[c(1:12, 14:26)]),
#               showRemainingTaxa = FALSE, 
#               plot_na = TRUE,
#               plot_values = FALSE,
#               plot_values_size = 3,
#               round = 2,
#               color_vector = c("white", colors[3]),
#               plot_legendbreaks = c(0.01, 0.1, 1, 10, 30),
#               plot_colorscale = "log10") +
#   scale_y_discrete(labels = rev(c(phylum.text.renamed[c(1:12, 14:26)]))) +
#   scale_x_discrete(labels = c("50 mg", "125 mg")) +
#   labs(x = "", 
#        y = "", 
#        tag = "",
#        title = "Heatmap aggregated at the phylum level\nacross DNA sample input") +
#   articletheme 
#   #theme(plot.margin = unit(c(0, 1, 0, 0), "cm"))

plot.heatmap_phylum <- dat.types.sub3000.amp %>%
  amp_subset_samples(normalise = TRUE) %>%
  amp_heatmap(group_by = "sample_target", 
              min_abundance = 0.001,
              facet_by = "soil_type",
              tax_aggregate = "Genus",
              #tax_add = c("Phylum"),
              normalise = FALSE,
              tax_show = 25,
              #order_y_by = "cluster",
              # tax_show = phylum.text.amp[c(1:12, 14:26)],
              # order_y_by = rev(phylum.text.amp[c(1:12, 14:26)]),
              showRemainingTaxa = FALSE,
              plot_na = TRUE,
              plot_values = FALSE,
              plot_values_size = 2,
              round = 1,
              #color_vector = c("white", colors[3]),
              plot_legendbreaks = c(0.01, 0.5, 2, 4, 6, 8),
              plot_colorscale = "sqrt") +
  guides(fill = guide_colourbar(ticks.colour = "black")) +
  scale_x_discrete(labels = c("50 mg", "125 mg")) +
  labs(x = "",
       y = "",
       tag = "",
       title = "Heatmap aggregated at the genus level\nacross DNA sample input") +
  articletheme +
  theme(strip.text.x = element_text(size = 7, face = "bold"))
  #theme(plot.margin = unit(c(0, 1, 0, 0), "cm"))

plot.heatmap_phylum

#leg.plot_heat <- cowplot::get_legend(plot.heatmap_phylum)

plot.heatmap_phylum.final <- plot.heatmap_phylum + theme(legend.position = "none")


#ggsave(here("Bead_beating", "figures", "heatmap_beadbeating_scale.png"), plot = plot.heatmap_phylum, device = "png", height = 12, width = 19)
```


# PERMANOVA test for overall difference in observed communities based on kit
```{r statistical testing}
## Overall comparison
permanova.list <- list()
      
for (i in 1:length(soil)) {
  
  community_0.1 <- dat.types.sub3000.amp %>%
    amp_subset_samples(normalise = TRUE) %>%
    filter_otus(0.1) %>%
    amp_subset_samples(soil_type == soil[i])
  
  community_0.1 <- t(community_0.1[["abund"]])
  
  meta <- dat.types.sub3000.amp[["metadata"]] %>%
    filter(soil_type == soil[i])
  
  dis <- vegan::vegdist(community_0.1)
  
  groups.stratified <- meta %>%
    mutate(group = str_c(soil_type, kit, sep = " ")) %>%
    pull(group) %>%
    as.factor(.)
  
 permanova <- adonis2(community_0.1 ~ sample_target, type = "bray", data = meta, permutations = 99999) %>%
   as.data.frame() %>%
   rownames_to_column("Variable") %>%
   mutate(across(R2, ~ as.character(round(. * 100, digits = 1)))) %>%
   mutate(pval = ifelse(`Pr(>F)` < 0.001, "< 0.001", round(`Pr(>F)`, 2))) %>%
   select(Variable, R2, pval)
 
 permanova.list[[i]] <- permanova
}
  
names(permanova.list) <- soil

perm.results <- as.numeric(c(permanova.list[[1]][1,2], permanova.list[[2]][1,2], permanova.list[[3]][1,2]))

shapiro.test(perm.results)
mean(perm.results)
sd(perm.results)

# Statified PERMANOVA
# Betadispersion - analysis of multivariate homogeneity of group dispersions (variances)
community_0.1 <- dat.types.sub3000.amp %>%
  amp_subset_samples(normalise = TRUE) %>%
  filter_otus(0.1)

community_0.1 <- t(community_0.1[["abund"]])

meta <- dat.types.sub3000.amp[["metadata"]]

dis <- vegan::vegdist(community_0.1)

groups.stratified <- meta %>%
  mutate(group = str_c(soil_type, sample_target, sep = " ")) %>%
  pull(group) %>%
  as.factor(.)

mod.stratified <- vegan::betadisper(dis, groups.stratified)
mod.stratified

anova(mod.stratified)

plot(mod.stratified, ellipse = TRUE, hull = FALSE, conf = 0.90) # 90% data ellipse

perm.strat <- with(meta, permute::how(nperm = 9999, blocks = soil_type))

set.seed(123)
permanova.target <- adonis2(community_0.1 ~ sample_target*soil_type, type = "bray", data = meta, permutations = 99999)

permanova.target %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(across(R2, ~ as.character(round(. * 100, digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 3))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

#write_csv(bray_distance, here("Bead_betaing", "figures", "2023-03-04_bray-distance.csv"))
#write_csv(comp.bray, here("Bead_beating", "figures", "2023-03-04_bray-comparisons.csv"))
```


## Ordinations - faceted by soil types
```{r ordination}
plot.ordinate <- dat.types.sub3000.amp %>%
  amp_ordinate(
    filter_species = 0.1,
    #sample_point_size = 5,
    constrain = "rpm",
    sample_shape_by = "sample_target",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 8,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = soil_type), alpha = 0.5, color = "black") +
  #scale_color_manual(values = colors) +
  scale_color_manual(values = colors[c(1,3,4,5,2)]) +
  scale_fill_manual(values = colors[c(1,3,4,5,2)]) +
  scale_shape_manual(values = c(16,21), labels = c("50 mg", "125 mg")) +
  labs(title = "All samples: PCA of microbial diversity",
       fill = "Soil type",
       shape = "Sample input [mg]",
       tag = "A",
       subtitle = paste0("**PERMANOVA:**", 
                         "<br>Soil type: Variance explained = ","89.5 %,", " p-value < 0.001",
                         "<br>Input amount: Variance explained = ","0.7 %,", " p-value = 0.14",
                         "<br>Intercation: Variance explained = ","2.0 %,", " p-value = 0.25")) +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8)) +
  articlethemex0 +
  theme(legend.position = "right",
        plot.subtitle = ggtext::element_markdown())

plot.ordinate

plot_ord1 <- dat.types.sub3000.amp %>%
  amp_subset_samples(soil_type == soil[1]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "sample_target",
    constrain = "rpm",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = sample_target), alpha = 0.5, color = "black") +
  scale_fill_manual(values = colors[1]) +
  scale_shape_manual(values = c(16,21), labels = c("50 mg", "125 mg")) +
  labs(title = paste0(soil[1], ": ", "PCA of microbial diversity"),
       tag = "B",
       subtitle = paste0("**PERMANOVA:**", 
                         "<br>Input amount: Variance explained = ", permanova.list[[1]][1,2], " %,", " p-value ", permanova.list[[1]][1,3],
                         "<br>",
                         "<br>")) +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8)) +
  articlethemex0 +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())
  
plot_ord2 <- dat.types.sub3000.amp %>%
  amp_subset_samples(soil_type == soil[2]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "sample_target",
    constrain = "rpm",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = sample_target), alpha = 0.5, color = "black") +
  scale_fill_manual(values = colors[3]) +
  scale_shape_manual(values = c(16,21), labels = c("50 mg", "125 mg")) +
  labs(title = paste0(soil[2], ": ", "PCA of microbial diversity"),
       tag = "C",
       subtitle = paste0("**PERMANOVA:**", 
                         "<br>Input amount: Variance explained = ", permanova.list[[2]][1,2], " %,", " p-value ", permanova.list[[2]][1,3],
                         "<br>",
                         "<br>")) +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8)) +
  articlethemex0 +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())

plot_ord3 <- dat.types.sub3000.amp %>%
  amp_subset_samples(soil_type == soil[3]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "sample_target",
    constrain = "rpm",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = sample_target), alpha = 0.5, color = "black") +
  scale_fill_manual(values = colors[4]) +
  scale_shape_manual(values = c(16,21), labels = c("50 mg", "125 mg")) +
  labs(title = paste0(soil[3], ": ", "PCA of microbial diversity"),
       tag = "D",
       subtitle = paste0("**PERMANOVA:**", 
                         "<br>Input amount: Variance explained = ", permanova.list[[3]][1,2], " %,", " p-value ", permanova.list[[3]][1,3],
                         "<br>",
                         "<br>")) +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8)) +
  articlethemex0 +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())

plot_ord4 <- dat.types.sub3000.amp %>%
  amp_subset_samples(soil_type == soil[4]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "sample_target",
    constrain = "rpm",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = sample_target), alpha = 0.5, color = "black") +
  scale_fill_manual(values = colors[5]) +
  scale_shape_manual(values = c(16,21), labels = c("50 mg", "125 mg")) +
  labs(title = paste0(soil[4], ": ", "PCA of microbial diversity"),
       tag = "E",
       subtitle = paste0("**PERMANOVA:**", 
                         "<br>Input amount: Variance explained = ", permanova.list[[4]][1,2], " %,", " p-value ", permanova.list[[4]][1,3],
                         "<br>",
                         "<br>")) +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8)) +
  articlethemex0 +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())

plot_ord5 <- dat.types.sub3000.amp %>%
  amp_subset_samples(soil_type == soil[5]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "sample_target",
    constrain = "rpm",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = sample_target), alpha = 0.5, color = "black") +
  scale_fill_manual(values = colors[2]) +
  scale_shape_manual(values = c(16,21), labels = c("50 mg", "125 mg")) +
  labs(title = paste0(soil[5], ": ", "PCA of microbial diversity"),
       tag = "F",
       subtitle = paste0("**PERMANOVA:**", 
                         "<br>Input amount: Variance explained = ", permanova.list[[5]][1,2], " %,", " p-value ", permanova.list[[5]][1,3],
                         "<br>",
                         "<br>")) +
  coord_cartesian(xlim = c(-0.8, 0.8), ylim = c(-0.8, 0.8)) +
  articlethemex0 +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())

leg.plot.ord_shape <- cowplot::get_legend(plot.ordinate + guides(fill = "none"))

leg.plot.ord_fill <- cowplot::get_legend(plot.ordinate + guides(shape = "none"))

plot_ord_leg <- cowplot::plot_grid(leg.plot.ord_fill, leg.plot.ord_shape, ncol = 1, align = "v", nrow = 2, rel_heights = c(1, 1))

plot.ord <- ggarrange(plot.ordinate, plot_ord1, plot_ord2, plot_ord3, plot_ord4, plot_ord5, common.legend = TRUE, ncol = 3, nrow = 2, legend = "top")

plot.ord

#plot_ord_final <- plot_ord3 +
#  labs(fill = "Soil type",
#       tag = "",
#       shape = "Library volume", 
#       subtitle = paste0("**PERMANOVA:**", "<br>Library volume: Variance explained = ", permanova.list[[3]][1,2], " %,", " p-value = ", permanova.list[[3]][1,3])) +
#  theme(legend.position = "none")

ggsave(here("Bead_beating", "figures", "pre_subfile2_Fig3.png"), plot = plot.ord, device = "png", height = 12, width = 19)
ggsave(here("Bead_beating", "figures", "pre_subfile2_Fig3.tif"), plot = plot.ord, device = "tiff", height = 12, width = 19, dpi = 600)
```


## Rel abundance correlation plot
# Next chunk merges replicates and sample type. It needs to be one plot per soil type
```{r differential relative abundance}
deseq_results <- tibble(soil = soil) %>%
  mutate(deseq = map(.x = soil, .f = ~ deseq_function_scale(.x, dat.types.sub3000.amp)))

deseq_results_dataframe <- rlist::list.rbind(deseq_results$deseq) %>%
  mutate(pval_signif = if_else(padj < 0.05, TRUE, FALSE),
         fc_signif = if_else(abs(log2FoldChange) > 1, TRUE, FALSE),)

dat_types_filtered <- dat.types.sub3000.amp %>%
  amp_subset_samples(normalise = TRUE)
  #filter_otus(0.1)

otu <- dat_types_filtered[["abund"]]

ProtocolDifferentialAbundance <- combine_otu_w_metadata_scale(metadata = dat.types.sub3000.amp$metadata, otu = otu) %>%
  summarise_triplicates_scale() %>%
  #calculate_relative_abundance_scale() %>%
  filter_double_zeros_scale() %>%
  left_join(deseq_results_dataframe) %>%
  mutate(Bias = if_else(log2FoldChange > 0 & pval_signif, "More abundant with 125 mg", 
                        if_else(pval_signif & log2FoldChange < 0, "More abundant with 50 mg", "Equally detected")),
         max_abundance = pmax(`50`, `125`)) %>%
  filter(!is.na(Bias))

points.ns <- ProtocolDifferentialAbundance %>%
  filter(Bias == "Equally detected")

points.signf <- ProtocolDifferentialAbundance %>%
  filter(!Bias == "Equally detected")

plot.diff <- ggplot() +
  geom_point(data = points.ns, aes(x = `125`, y = `50`, fill = Bias), alpha = 0.2, 
             color = "black", size = 4, position = position_jitter(width = 0.001), shape = 21) +
  geom_point(data = points.signf, aes(x = `125`, y = `50`, fill = Bias), alpha = 0.8, 
             color = "black", size = 4, position = position_jitter(width = 0.001), shape = 21) +
  coord_cartesian(xlim = c(0, max(ProtocolDifferentialAbundance$max_abundance)), ylim = c(0, max(ProtocolDifferentialAbundance$max_abundance))) +
  scale_fill_manual(values = c("grey", colors[5], colors[3]), 
                    limits = c("Equally detected",
                               "More abundant with 125 mg", 
                               "More abundant with 50 mg")) +
  geom_abline(linewidth = 0.5) +
  articlethemex0 +
    labs(title = "All samples: Differential abundance of ASVs",
         x = "Average relative abundance within replicates [%], 125 mg",
         y = "Average relative abundance within replicates [%], 50 mg"
    )

plot.diff

plot_list <- map(.x = soil, ~ plotDiffRelabund_scale(ProtocolDifferentialAbundance, .x)) 

ggleg <- plot_list[[3]] + theme(legend.position = "right") 

leg <- list(get_legend(ggleg))

plot_diff <- plot.diff + 
  labs(tag = "B") + 
  #labs(title = "Organic: Differential abundance of ASVs", subtitle =  "ASVs with relative abundance above 0.1 %", tag = "B") + 
  #coord_cartesian(xlim = c(0, 0.2), ylim = c(0, 0.2)) +
  theme(legend.position = "none", plot.subtitle = element_text(hjust = 0.5))

plot_list2 <- append(plot_list, list(plot.diff), after = 0)

p.diff <- ggarrange(plotlist = plot_list2, common.legend = TRUE, ncol = 3, nrow = 2, legend = "top", labels = toupper(letters[1:6]), font.label = list(size = 18, face = "bold"))

p.diff
```

Note on p-values set to NA: some values in the results table can be set to NA for one of the following reasons:

* If within a row, all samples have zero counts, the baseMean column will be zero, and the log2 fold change estimates, p value and adjusted p value will all be set to NA.
* If a row contains a sample with an extreme count outlier then the p value and adjusted p value will be set to NA. These outlier counts are detected by Cook’s distance. Customization of this outlier filtering and description of functionality for replacement of outlier counts and refitting is described below
* If a row is filtered by automatic independent filtering, for having a low mean normalized count, then only the adjusted p value will be set to NA. Description and customization of independent filtering is described below
`NAs can be omitted by setting independentFiltering = FALSE, cooksCutoff = FALSE`

## Final plot

```{r final plot}
# Create combined plot
leg.heat <- cowplot::get_legend(plot.heatmap_phylum)
leg.ord <- cowplot::get_legend(ggleg)

leg.comb <- cowplot::plot_grid(leg.heat, leg.ord, ncol = 1, align = "v", nrow = 2, rel_heights = c(1, 1))

leg.comb

p.final <- ggarrange(plot.heatmap_phylum.final, plot_list[[1]], plot_list[[2]], plot_list[[3]], plot_list[[4]], plot_list[[5]], 
                     ncol = 3, nrow = 2, widths = c(1, 1, 1), labels = toupper(letters[1:6]), font.label = list(size = 18, face = "bold"))

p.final

gg_arranged <- ggarrange(p.final, leg.comb, ncol = 2, nrow = 1, widths = c(1, 0.15))

gg_arranged

ggsave(here("Bead_beating", "figures", "pre_subfile2_Fig2.png"), plot = gg_arranged, device = "png", height = 12, width = 20)
ggsave(here("Bead_beating", "figures", "pre_subfile2_Fig2.tif"), plot = gg_arranged, device = "tiff", height = 12, width = 20, dpi = 600)
```









