---
title: "Amplicon diversity of MFD standard samples in extraction benchmark 2.0"
author: "Thomas Bygh Nymann Jensen"
date: "`r format(Sys.time(), '%d-%m-%Y')`, Aalborg, Denmark"
output: 
  bookdown::html_document2:
    toc: true
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo    = F,
  # fig.pos = "H",
  message = FALSE,
  warning = FALSE
)

library(here)
setwd(here("Benchmarking"))

# Load the needed packages.
#library(data.table)
#library(readxl)
#library(writexl)
library(tidyverse)
library(ampvis2)
library(DESeq2)
library(vegan)
library(patchwork)
library(ggpubr)
#library(ggrepel)

### Load and tidy ###
source(here("Benchmarking", "scripts/loaddata.R"))
source(here("Benchmarking", "scripts/functions.R"))
source(here("Benchmarking", "scripts/article_theme.R"))

colors <- c("grey", "darkgoldenrod2", "royalblue4", "seagreen", "salmon")
kit_colors <- RColorBrewer::brewer.pal(5, "Dark2")
```


## Filter Data

> Sample MG200131-87 is a clear outlier in PCA space. Therefore it is removed. # NOT REMOVED!

```{r remove blanks and pcr controls}
dat.types.amp <- ampvis.amplicon %>%
  amp_subset_samples(!is.na(project_id) & !soil_type %in% c("Blank", "Activated sludge")) %>%
  amp_subset_samples(kit != "MIDAS HT")#, seq_name != "MG200131-87")

# Create subsets
dat.types.sub10000.amp <- dat.types.amp %>% amp_subset_samples(., minreads = 10000)

rarefy <- min(colSums(dat.types.sub10000.amp[["abund"]]))

dat.types.sub10000.amp.ra <- dat.types.sub10000.amp %>%
  amp_subset_samples(., rarefy = rarefy)

dat.types.sub10000.amp.rel_0.1 <- dat.types.sub10000.amp %>%
  amp_subset_samples(normalise = TRUE) %>%
  filter_otus(0.1)
```

We set a cutoff of at 10000 reads, and rarefy to the lowest number of reads in the remaining dataset. Doing this we go from `r nrow(dat_types.amp$metadata)` to `r nrow(dat.types.sub10000.amp$metadata)` samples and create a rarified dataset as well.

### Bray Curtis variation
# Only OTUs above 0.1 % relative abundance is used

>MIDAS was removed at this stage.

```{r}
dat.types.sub10000.amp %>%
  amp_rarecurve(color_by = "kit",
                facet_by = "soil_type",
                facet_scales = "free_x",
                stepsize = 1000) +
  scale_color_manual(values = c(kit_colors))

bray_distance <- as.matrix(vegan::vegdist(t(dat.types.sub10000.amp.rel_0.1$abund))) %>%
    as.data.frame() %>%
    rownames_to_column("seq_id")
  
metadata <- dat.types.sub10000.amp.rel_0.1$metadata %>%
  select(seq_id, kit, soil_type)

metadata_renamed <- metadata %>%
  dplyr::rename(
    comparison = seq_id,
    compared_soil = soil_type,
    compared_kit = kit
  )

bray_with_metadata <- metadata %>%
  left_join(bray_distance) %>%
  pivot_longer(!seq_id:soil_type, names_to = "comparison", values_to = "value") %>%
  left_join(metadata_renamed)
  
bray_with_metadata_comparisons <- bray_with_metadata %>%
  filter(seq_id != comparison) %>%
  filter(paste0(kit, soil_type) == paste0(compared_kit, compared_soil)) %>%
  filter(!duplicated(paste0(pmax(seq_id, comparison), pmin(seq_id, comparison)))) %>%
  group_by(soil_type, kit) %>%
  summarise(
    mean_bray = format(round(mean(value, na.rm = TRUE), 2), nsmall = 2),
    sd_bray = format(round(sd(value, na.rm = TRUE), 2), nsmall = 2)
  )
```


### Shannon Diversity variation
```{r}
tab_reads <- data.frame(num_reads = apply(dat.types.sub10000.amp$abund, 2, sum)) %>%
  rownames_to_column(var = "seq_id")

lab_metadata <- metadata.amplicon %>%
  select(seq_id, kit, soil_type, conc_qubit, `260/280`, `260/230`, lib_conc) %>%
  filter(kit != "MIDAS HT", !soil_type %in% c("Blank", "Activated sludge")) %>%
  mutate(yield_ng = if_else(kit == "PowerSoil Pro HT", conc_qubit * 120, conc_qubit * 60))

rarefied_alphadiversity <- dat.types.sub10000.amp.ra %>%
  amp_alphadiv(
    measure = c("observed", "shannon", "simpson"),
    richness = T
  ) %>%
  dplyr::rename(shannon = Shannon,
                observed_asv = ObservedOTUs) %>%
  select(
    seq_id,
    soil_type,
    kit,
    observed_asv,
    shannon
  ) %>%
  left_join(tab_reads)

# Join different data: Lab, Alpha, beta.
combine_alpha_beta_with_metadata <- rarefied_alphadiversity %>%
  left_join(lab_metadata) %>%
  # Calculate failed libraries - FastSpin for sand worked however one sample was an outlier
  group_by(soil_type, kit) %>%
  mutate(libs = sum(!is.na(shannon))) %>%
  # Calculate mean and sd for all variables
  summarise(
    across(where(is.numeric), .fns = list(mean = ~ mean(., na.rm = T), sd = ~ sd(., na.rm = T)), .names = "{.fn}_{.col}")
  ) %>%
  # Format mean and sd
  ungroup() %>%
  mutate(
    libs = as.character(mean_libs),
    across(mean_num_reads:sd_num_reads, ~ format(round(., 0))),
    across(mean_observed_asv:sd_observed_asv, ~ format(round(., 0))),
    across(where(is.numeric), ~ format(round(., 2), nsmall = 2)),
    mean_yield_ng = format(round(as.numeric(mean_yield_ng))),
    sd_yield_ng = format(round(as.numeric(sd_yield_ng)))) %>%
  # Add bray variation
  left_join(bray_with_metadata_comparisons) %>%
  # factor Kit
  mutate(kit = factor(kit, levels = kit_levels, labels = kit_levels)) %>%
  arrange(soil_type, kit)
```


## Final table
```{r}
mean_sd_table_long <- combine_alpha_beta_with_metadata %>%
  select(-c(sd_libs, mean_libs)) %>%
  pivot_longer(cols = -c(soil_type, kit, libs, contains("sd")), names_to = "mean", values_to = "mean_value", names_prefix = "mean_") %>%
  pivot_longer(cols = -c(soil_type, kit, libs, contains("mean")), names_to = "sd", values_to = "sd_value", names_prefix = "sd_") %>%
  group_by(soil_type, kit) %>%
  filter(mean == sd) %>%
  mutate(
    mean_sd = paste0(mean_value, " (", sd_value, ")")
  ) %>%
  select(!mean_value:sd_value)

mean_sd_table <- mean_sd_table_long %>%
  pivot_wider(names_from = mean, values_from = mean_sd) %>%
  select(soil_type, kit, conc_qubit, yield_ng, `260/280`, `260/230`, libs, lib_conc, num_reads, observed_asv, shannon, bray) %>%
  group_by(soil_type) %>%
  mutate_at(vars(soil_type), ~ replace(., duplicated(.), ""))

names <- tibble(
  old_names = colnames(mean_sd_table),
  new_names = c("Soil type", "Kit", "DNA ext conc. [ng/µL]", "DNA yield [ng]", "260/280", "260/230", "Libraries", "Library conc. [ng/µL]", "Number of reads", "Observed ASVs", "Shannon Diversity", "Bray-Curtis Variation") # "Site name", "Longitude Latitude", "Environment",
)

final_table <- mean_sd_table %>%
  rename_at(vars(names$old_names), ~ names$new_names)

hjustification_matrix <- as.vector(matrix(c(0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1), ncol = ncol(mean_sd_table), nrow = nrow(mean_sd_table), byrow = TRUE))

xjustification_matrix <- as.vector(matrix(c(0.1, 0.1, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9), ncol = ncol(mean_sd_table), nrow = nrow(mean_sd_table), byrow = TRUE))

alpha_table <- final_table %>%
  ggtexttable(rows = NULL, theme = ttheme(colnames.style = colnames_style(fill = "white"), tbody.style = tbody_style(fill = "white", hjust = hjustification_matrix, x = xjustification_matrix))) %>%
  tab_add_hline(at.row = 1:2, row.side = "top", linewidth = 2) %>%
  tab_add_hline(at.row = seq(7, 24, 5), row.side = "top", linewidth = 1)

alpha_table

ggsave(here("Benchmarking", "figures", "table_benchmarking.png"), plot = alpha_table, device = "png", height = 12, width = 19)

writexl::write_xlsx(final_table, here("Benchmarking", "figures", "benchmark_overview_table.xlsx"))
```


## Statistical tests - two sample t-test or Mann-Whitney(Wilcoxon) on shannon and bray distance
```{r}
alpha.all <- rarefied_alphadiversity %>%
  right_join(lab_metadata) 

alpha.ht <- alpha.all %>%
  filter(str_detect(kit, "HT"))

comp.bray <- bray_with_metadata %>%
  filter(value != 0) %>%
  filter(soil_type == compared_soil, kit == compared_kit)

# ANOVA yield
anova(lm(yield_ng ~ kit * soil_type, data = alpha.ht)) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ round(., digits = 2))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "<0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

# ANOVA shannon
anova(lm(shannon ~ kit * soil_type, data = alpha.ht)) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ round(., digits = 2))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "<0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

# ANOVA bray
anova(lm(value ~ kit * soil_type, data = comp.bray)) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ round(., digits = 2))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "<0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)


yield.FS_HT <- alpha.all %>% filter(kit == "FastSpin HT") %>% pull(yield_ng)

yield.ZM_HT <- alpha.all %>% filter(kit == "ZymoMagbead HT") %>% pull(yield_ng)

yield.PS_HT <- alpha.all %>% filter(kit == "PowerSoil Pro HT") %>% pull(yield_ng)

test.yield.FS_PS <- wilcox.test(yield.FS_HT, yield.PS_HT, alternative = "less")
test.yield.ZM_PS <- wilcox.test(yield.ZM_HT, yield.PS_HT, alternative = "less")
test.yield.FS_ZM <- wilcox.test(yield.FS_HT, yield.ZM_HT, alternative = "less")

p.yield <- c(test.yield.FS_PS$p.value, test.yield.ZM_PS$p.value, test.yield.FS_ZM$p.value)
p.adj_yield <- p.adjust(p.yield, method = "bonferroni")
p.adj_yield

ratio1.PS_LT <- alpha.all %>% filter(kit == "PowerSoil LT") %>% pull(`260/280`)
ratio1.PS_HT <- alpha.all %>% filter(kit == "PowerSoil Pro HT") %>% pull(`260/280`)
ratio2.PS_LT <- alpha.all %>% filter(kit == "PowerSoil LT") %>% pull(`260/230`)
ratio2.PS_HT <- alpha.all %>% filter(kit == "PowerSoil Pro HT") %>% pull(`260/230`)

test.ratio1 <- wilcox.test(ratio1.PS_LT, ratio1.PS_HT, alternative = "less")
test.ratio2 <- wilcox.test(ratio2.PS_LT, ratio2.PS_HT, alternative = "less")

p.ratio <- c(test.ratio1$p.value, test.ratio2$p.value)
p.adj_ratio <- p.adjust(p.ratio, method = "bonferroni")
p.adj_ratio
```


## Heatmap
```{r}
# Find all phyla in amplicon data
heatmap.amplicon_phylum <- dat.types.sub10000.amp %>%
  amp_heatmap(group_by = "seq_id", 
              min_abundance = 0.001,
              tax_show = 1000,
              tax_aggregate = "Phylum", 
              normalise = FALSE,
              showRemainingTaxa = FALSE, 
              plot_values = TRUE,
              textmap = TRUE) %>%
  rownames_to_column(var = "Phylum") %>%
  mutate(across(Phylum, ~ str_replace_all(., "^.*ASV.*$", "Unclassified"))) %>%
  group_by(Phylum) %>%
  summarise(across(everything(), ~ sum(.))) %>%
  ungroup() %>%
  column_to_rownames(var = "Phylum")

phylum.text.amp <- heatmap.amplicon_phylum %>%
  mutate(total = rowSums(.)) %>%
  filter(!total == 0) %>%
  select(total) %>%
  arrange(desc(total)) %>%
  row.names(.)

phylum.text.renamed <- gsub("SAR324_cladeMarine_group_B", "SAR324", phylum.text.amp)

# Load new genus ampvis object
ampvis.sub_phylum <- amp_load(otutable = heatmap.amplicon_phylum,
                              metadata = metadata.amplicon)

plot.heatmap_phylum <- ampvis.sub_phylum %>%
  amp_heatmap(group_by = "kit", 
              min_abundance = 0.01,
              facet_by = "soil_type",
              tax_aggregate = "OTU",
              normalise = TRUE,
              tax_show = phylum.text.amp[c(1:14, 16:26)], 
              order_y_by = rev(phylum.text.amp[c(1:14, 16:26)]),
              showRemainingTaxa = FALSE, 
              plot_na = TRUE,
              plot_values = FALSE,
              plot_values_size = 3,
              round = 2,
              color_vector = c(colors[5], "white", colors[3]),
              plot_legendbreaks = c(0.01, 0.1, 1, 10, 30),
              plot_colorscale = "log10") +
  scale_y_discrete(labels = rev(c(phylum.text.renamed[c(1:14, 16:26)]))) +
  labs(x = "", y = "", title = "Heatmap aggregated at the phylum level across extraction kits") +
  articletheme +
  guides(fill = guide_colorbar(title.position = "top"))

leg.plot_heat <- cowplot::get_legend(plot.heatmap_phylum)

plot.heatmap_phylum.final <- plot.heatmap_phylum + theme(legend.position = "none")

plot.heatmap_phylum

ggsave(here("Benchmarking", "figures", "heatmap_benchmarking.png"), plot = plot.heatmap_phylum, device = "png", height = 12, width = 19)
```


## Ordinations - faceted by soil types
```{r ordination}
soil <- unique(dat.types.sub10000.amp$metadata$soil_type)

plot_soil_ord <- function(soil) {
  plotAmpliconOrdinationSoil(dat.types.sub10000.amp, soil, ord_type = "PCA") +
    scale_color_manual(values = c(kit_colors)) +
    scale_fill_manual(values = c(kit_colors))
}

plot_list <- map(soil, plot_soil_ord)

ggleg <- plot_list[[1]] + theme(legend.position = "right")

leg <- list(get_legend(ggleg))

leg.plot_ord <- cowplot::get_legend(ggleg)

plot_ord <- plot_list[[4]] + 
  coord_equal() +
  labs(fill = "Extraction kit",
       color = "Extraction kit") +
  theme(legend.position = "right")

leg.plot_ord <- cowplot::get_legend(plot_ord)

plot_ord_final <- plot_ord + theme(legend.position = "none")

plot_list2 <- append(plot_list, leg)

p.ord <- ggarrange(plotlist = plot_list2, nrow = 2, ncol = ceiling(length(plot_list) / 2), labels = toupper(letters[1:5]))

p.ord

ggsave(here("Benchmarking", "figures", "PCA_faceted_benchmarking.png"), plot = p.ord, device = "png", height = 12, width = 19)
```


## Ordination combined
```{r}
plot.ordinate <- dat.types.sub10000.amp %>%
  amp_ordinate(filter_species = 0.1,
               transform = "hellinger",
               distmeasure = "bray",
               sample_shape_by = "kit",
               #sample_colorframe = "soil_type",
               species_nlabels = 10,
               species_label_taxonomy = "Phylum",
               species_plot = T) +
  scale_color_manual(values = colors) +
  scale_fill_manual(values = colors) +
  labs(title = "PCA of microbial diversity",
       fill = "Sample type",
       shape = "Extraction kit") +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = soil_type), alpha = 0.5, color = "black") +
  coord_equal() +
  articletheme

plot.ordinate

#The plot with a zoom
plot.insert <- plot.ordinate +
  ylim (0.54, 0.57) +
  xlim (0.19, 0.22) +
  theme(plot.title = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none") +
  theme(axis.text.x = element_blank(),
        axis.text.y = element_blank())

plot.insert

plot.ordinate.insert <- plot.ordinate + 
  #geom_segment(aes(x = 0.19, y = 0.57, xend = -0.3, yend = 0.6), color = "black", linetype = "dashed") +
  geom_segment(aes(x = 0.22, y = 0.57, xend = 0.1, yend = 0.6), color = "black", linetype = "dashed") +
  #geom_segment(aes(x = 0.19, y = 0.54, xend = -0.3, yend = 0.25), color = "black", linetype = "dashed") +
  geom_segment(aes(x = 0.22, y = 0.54, xend = 0.1, yend = 0.25), color = "black", linetype = "dashed") +
  annotation_custom(ggplotGrob(plot.insert), xmin = -0.3, xmax = 0.1, ymin = 0.25, ymax = 0.6) +
  geom_rect(aes(xmin = 0.19, xmax = 0.22, ymin = 0.54, ymax = 0.57), color='black', alpha=0) +
  geom_rect(aes(xmin = -0.3, xmax = 0.1, ymin = 0.25, ymax = 0.6), color = "black", linetype = "dashed", alpha = 0)

plot.ordinate.insert

leg.plot_ord.insert <- cowplot::get_legend(plot.ordinate.insert)

plot.ordinate.insert_final <- plot.ordinate.insert + theme(legend.position = "none")

ggsave(here("Benchmarking", "figures", "PCA_ordination_insert.png"), plot = plot.ordinate.insert, device = "png", height = 12, width = 19)
```


# PERMANOVA test for overall difference in observed communities based on kit
```{r statistical testing}
## Overall comparison
# Betadispersion - analysis of multivariate homogeneity of group dispersions (variances)
community <- dat.types.sub10000.amp.rel_0.1

community_0.1 <- t(community[["abund"]])

meta <- dat.types.sub10000.amp.rel_0.1[["metadata"]]

dis <- vegan::vegdist(community_0.1)

groups.stratified <- meta %>%
  mutate(group = str_c(soil_type, kit, sep = " ")) %>%
  pull(group) %>%
  as.factor(.)

mod.stratified <- vegan::betadisper(dis, groups.stratified)
mod.stratified

anova(mod.stratified)

plot(mod.stratified, ellipse = TRUE, hull = FALSE, conf = 0.90) # 90% data ellipse

perm.strat <- with(meta, permute::how(nperm = 9999, blocks = soil_type))

set.seed(123)
permanova.stratified <- adonis2(community_0.1 ~ kit * soil_type, type = "bray", data = meta, permutations = perm.strat)

permanova.stratified %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(across(R2, ~ round(., digits = 2))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "<0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)
```

## Final plot
```{r final plot}
# Create combined plot
gg1 <- ggarrange(plot.heatmap_phylum.final, labels = c("A"))

gg2 <- ggarrange(plot.ordinate.insert_final, plot_ord_final, labels = c("B", "C"), ncol = 1)

gg3 <- ggarrange(leg.plot_heat, leg.plot_ord, ncol = 1)

gg4 <- ggarrange(gg3, leg.plot_ord.insert, ncol = 1)

gg_arranged <- ggarrange(gg1, gg2, gg4, ncol = 3, nrow = 1, widths = c(2, 1, 0.5))

gg_arranged

ggsave(here("Benchmarking", "figures", "combined_plot_benchmarking.png"), plot = gg_arranged, device = "png", height = 12, width = 19)
```





