---
title: "Amplicon diversity of MFD standard samples in extraction benchmark 2.0"
author: "Thomas Bygh Nymann Jensen"
date: "`r format(Sys.time(), '%d-%m-%Y')`, Aalborg, Denmark"
output: 
  bookdown::html_document2:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = F,
  # fig.pos = "H",
  message = FALSE,
  warning = FALSE
)

library(here)
setwd(here("Benchmarking"))

# Load the needed packages.
#library(data.table)
#library(readxl)
#library(writexl)
library(tidyverse)
library(ampvis2)
library(DESeq2)
library(vegan)
library(patchwork)
library(ggpubr)
#library(ggrepel)

### Load and tidy ###
source(here("Benchmarking", "scripts/loaddata.R"))
source(here("Benchmarking", "scripts/functions.R"))
source(here("Benchmarking", "scripts/article_theme.R"))

colors <- c("grey", "darkgoldenrod2", "royalblue4", "seagreen", "salmon")
#kit_colors <- RColorBrewer::brewer.pal(5, "Dark2")
```


## Filter Data

> Samples MQ200131-87 and MQ200131-66 are outliers in PCA space. Therefore they are removed.

```{r remove blanks and pcr controls}
dat.types.amp <- ampvis.amplicon %>%
  amp_subset_samples(!is.na(project_id) & !soil_type %in% c("Blank", "Activated sludge")) %>%
  amp_subset_samples(kit != "MIDAS HT", !seq_name %in% c("MQ200131-66", "MQ200131-87"))

# Create subsets
dat.types.sub10000.amp <- dat.types.amp %>% amp_subset_samples(., minreads = 10000)

rarefy <- min(colSums(dat.types.sub10000.amp[["abund"]]))

set.seed(123)
dat.types.sub10000.amp.ra <- dat.types.sub10000.amp %>%
  amp_subset_samples(., rarefy = rarefy)

dat.types.sub10000.amp.ra.rel_0.1 <- dat.types.sub10000.amp %>%
  amp_subset_samples(normalise = TRUE) %>%
  filter_otus(0.1)

soil <- unique(dat.types.sub10000.amp$metadata$soil_type) %>% sort(.)
```

We set a cutoff of at 10000 reads, and rarefy to the lowest number of reads in the remaining dataset. Doing this we go from `r nrow(dat_types.amp$metadata)` to `r nrow(dat.types.sub10000.amp$metadata)` samples and create a rarified dataset as well.

### Bray Curtis variation
# Only OTUs above 0.1 % relative abundance is used

>MIDAS was removed at this stage.

```{r}
amp.rarecurve <- dat.types.sub10000.amp %>%
  amp_rarecurve(color_by = "kit",
                facet_by = "soil_type",
                facet_scales = "free_x",
                stepsize = 1000) +
  scale_color_manual(values = c(colors))

tmp <- t(dat.types.sub10000.amp.ra.rel_0.1$abund)

bray_distance <- as.matrix(vegan::vegdist(vegan::decostand(tmp, method = "hellinger"))) %>%
    as.data.frame() %>%
    rownames_to_column("seq_id")

#bray_distance <- as.matrix(vegan::vegdist(t(dat.types.sub10000.amp.ra.rel_0.1$abund))) %>%
#    as.data.frame() %>%
#    rownames_to_column("seq_id")
  
metadata <- dat.types.sub10000.amp.ra.rel_0.1$metadata %>%
  select(seq_id, kit, soil_type)

metadata_renamed <- metadata %>%
  dplyr::rename(
    comparison = seq_id,
    compared_soil = soil_type,
    compared_kit = kit
  )

bray_with_metadata <- metadata %>%
  left_join(bray_distance) %>%
  pivot_longer(!seq_id:soil_type, names_to = "comparison", values_to = "value") %>%
  left_join(metadata_renamed) %>%
  filter(value != 0) %>%
  distinct(value, .keep_all = TRUE)
  
bray_with_metadata_comparisons <- bray_with_metadata %>%
  filter(seq_id != comparison) %>%
  filter(paste0(kit, soil_type) == paste0(compared_kit, compared_soil)) %>%
  filter(!duplicated(paste0(pmax(seq_id, comparison), pmin(seq_id, comparison)))) %>%
  group_by(soil_type, kit) %>%
  summarise(
    mean_bray = format(round(mean(value, na.rm = TRUE), 2), nsmall = 2),
    sd_bray = format(round(sd(value, na.rm = TRUE), 2), nsmall = 2)) %>%
  ungroup() %>%
  mutate(across(mean_bray:sd_bray, ~ as.numeric(.)))
```


### Shannon Diversity variation
```{r}
tab_reads <- data.frame(num_reads = apply(dat.types.sub10000.amp$abund, 2, sum)) %>%
  rownames_to_column(var = "seq_id")

lab_metadata <- metadata.amplicon %>%
  select(seq_id, kit, soil_type, conc_qubit, `260/280`, `260/230`, integrity, lib_conc) %>%
  filter(kit != "MIDAS HT", !soil_type %in% c("Blank", "Activated sludge")) %>%
  mutate(yield_ng = if_else(kit == "PowerSoil Pro HT", conc_qubit * 120, conc_qubit * 60),
         across(integrity, ~as.numeric(.)))

rarefied_alphadiversity <- dat.types.sub10000.amp.ra %>%
  amp_alphadiv(
    measure = c("observed", "shannon", "simpson"),
    richness = T
  ) %>%
  dplyr::rename(shannon = Shannon,
                observed_asv = ObservedOTUs) %>%
  select(seq_id, soil_type, kit, observed_asv, shannon) %>%
  left_join(tab_reads)

# Join different data: Lab, Alpha, beta.
combine_alpha_beta_with_metadata <- rarefied_alphadiversity %>%
  right_join(lab_metadata) %>%
  # Calculate failed libraries - FastSpin for sand worked however one sample was an outlier
  group_by(soil_type, kit) %>%
  mutate(libs = sum(!is.na(shannon)),
         across(integrity, ~as.numeric(.))) %>%
  # Calculate mean and sd for all variables
  summarise(
    across(where(is.numeric), .fns = list(mean = ~ mean(., na.rm = T), sd = ~ sd(., na.rm = T)), .names = "{.fn}_{.col}")
  ) %>%
  # Format mean and sd
  ungroup() %>%
  mutate(
    libs = as.character(mean_libs),
    across(mean_num_reads:sd_num_reads, ~ format(round(., 0))),
    across(mean_observed_asv:sd_observed_asv, ~ format(round(., 0))),
    across(mean_lib_conc:sd_lib_conc, ~ format(round(., 2))),
    across(mean_integrity:sd_integrity, ~ format(round(./1000, 2))),
    mean_yield_µg = format(round(as.numeric(mean_yield_ng)/1000, 2)),
    sd_yield_µg = format(round(as.numeric(sd_yield_ng)/1000, 2))) %>%
  select(-mean_yield_ng, -sd_yield_ng) %>%
  # Add bray variation
  left_join(bray_with_metadata_comparisons) %>%
  # factor Kit
  mutate(kit = factor(kit, levels = kit_levels, labels = kit_levels),
         across(where(is.numeric), ~ format(round(., 2), nsmall = 2))) %>%
  arrange(soil_type, kit)
```


## Final table
```{r}
mean_sd_table_long <- combine_alpha_beta_with_metadata %>%
  select(-c(sd_libs, mean_libs)) %>%
  pivot_longer(cols = -c(soil_type, kit, libs, contains("sd")), names_to = "mean", values_to = "mean_value", names_prefix = "mean_") %>%
  pivot_longer(cols = -c(soil_type, kit, libs, contains("mean")), names_to = "sd", values_to = "sd_value", names_prefix = "sd_") %>%
  group_by(soil_type, kit) %>%
  filter(mean == sd) %>%
  mutate(
    mean_sd = paste0(mean_value, " (", sd_value, ")")
  ) %>%
  select(!mean_value:sd_value)

mean_sd_table <- mean_sd_table_long %>%
  pivot_wider(names_from = mean, values_from = mean_sd) %>%
  select(soil_type, kit, yield_µg, `260/280`, `260/230`, integrity, libs, lib_conc, num_reads, observed_asv, shannon, bray) %>%
  group_by(soil_type) %>%
  mutate_at(vars(soil_type), ~ replace(., duplicated(.), ""))

names <- tibble(
  old_names = colnames(mean_sd_table),
  new_names = c("Soil type", "Extraction Kit", "DNA yield [µg]", "260/280", "260/230", "Integrity [kbp]", "Libraries", "Library conc. [ng/µL]", "Number of reads", "Observed ASVs", "Shannon Diversity", "Bray-Curtis Dissimilarity") # "Site name", "Longitude Latitude", "Environment",
)

final_table <- mean_sd_table %>%
  rename_at(vars(names$old_names), ~ names$new_names)

hjustification_matrix <- as.vector(matrix(c(0, 0, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1), ncol = ncol(mean_sd_table), nrow = nrow(mean_sd_table), byrow = TRUE))

xjustification_matrix <- as.vector(matrix(c(0.1, 0.1, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9, 0.9), ncol = ncol(mean_sd_table), nrow = nrow(mean_sd_table), byrow = TRUE))

alpha_table <- final_table %>%
  ggtexttable(rows = NULL, theme = ttheme(colnames.style = colnames_style(fill = "white"), tbody.style = tbody_style(fill = "white", hjust = hjustification_matrix, x = xjustification_matrix))) %>%
  tab_add_hline(at.row = 1:2, row.side = "top", linewidth = 2) %>%
  tab_add_hline(at.row = seq(7, 24, 5), row.side = "top", linewidth = 1)

alpha_table

#ggsave(here("Benchmarking", "figures", "table_benchmarking.png"), plot = alpha_table, device = "png", height = 12, width = 19)

writexl::write_xlsx(final_table, here("Benchmarking", "figures", "S1_Table.xlsx"))
```


## Statistical tests - ANOVA and Mann-Whitney
```{r}
alpha.all <- rarefied_alphadiversity %>%
  right_join(lab_metadata) %>%
  arrange(shannon) %>%
  mutate(shannon_rank = seq(1:length(shannon))) %>%
  arrange(yield_ng) %>%
  mutate(yield_ng_rank = seq(1:length(yield_ng))) %>%
  arrange(`260/280`) %>%
  mutate(`260/280_rank` = seq(1:length(`260/280`))) %>%
  arrange(`260/230`) %>%
  mutate(`260/230_rank` = seq(1:length(`260/230`))) %>%
  arrange(integrity) %>%
  mutate(integrity_rank = seq(1:length(integrity))) %>%
  mutate(across(soil_type:kit, ~as.factor(.)))
  
comp.bray <- bray_with_metadata %>%
  dplyr::rename(bray = value) %>%
  filter(bray != 0) %>%
  filter(soil_type == compared_soil, kit == compared_kit) %>%
  arrange(bray) %>%
  mutate(bray_rank = seq(1:length(bray))) %>%
  mutate(across(c(kit,soil_type,compared_kit, compared_soil), ~as.factor(.)))

# Test for normality
shapiro.test(alpha.all$shannon)
qqnorm(alpha.all$shannon)
qqline(y = alpha.all$shannon)

bartlett.test(shannon ~ kit, data = alpha.all)
fligner.test(shannon ~ kit, data = alpha.all)
car::leveneTest(shannon ~ kit, data = alpha.all)

bartlett.test(shannon ~ soil_type, data = alpha.all)
fligner.test(shannon ~ soil_type, data = alpha.all)
car::leveneTest(shannon ~ soil_type, data = alpha.all)

shapiro.test(alpha.all$yield_ng)
qqnorm(alpha.all$yield_ng)
qqline(y = alpha.all$yield_ng)

bartlett.test(yield_ng ~ kit, data = alpha.all)
fligner.test(yield_ng ~ kit, data = alpha.all)
car::leveneTest(yield_ng ~ kit, data = alpha.all)

bartlett.test(yield_ng ~ soil_type, data = alpha.all)
fligner.test(yield_ng ~ soil_type, data = alpha.all)
car::leveneTest(yield_ng ~ soil_type, data = alpha.all)

shapiro.test(alpha.all$`260/280`)
qqnorm(alpha.all$`260/280`)
qqline(y = alpha.all$`260/280`)

bartlett.test(`260/280` ~ kit, data = alpha.all)
fligner.test(`260/280` ~ kit, data = alpha.all)
car::leveneTest(`260/280` ~ kit, data = alpha.all)

bartlett.test(`260/280` ~ soil_type, data = alpha.all)
fligner.test(`260/280` ~ soil_type, data = alpha.all)
car::leveneTest(`260/280` ~ soil_type, data = alpha.all)

shapiro.test(alpha.all$`260/230`)
qqnorm(alpha.all$`260/230`)
qqline(y = alpha.all$`260/230`)

bartlett.test(`260/230` ~ kit, data = alpha.all)
fligner.test(`260/230` ~ kit, data = alpha.all)
car::leveneTest(`260/230` ~ kit, data = alpha.all)

bartlett.test(`260/230` ~ soil_type, data = alpha.all)
fligner.test(`260/230` ~ soil_type, data = alpha.all)
car::leveneTest(`260/230` ~ soil_type, data = alpha.all)

shapiro.test(alpha.all$integrity)
qqnorm(alpha.all$integrity)
qqline(y = alpha.all$integrity)

bartlett.test(integrity ~ kit, data = alpha.all)
fligner.test(integrity ~ kit, data = alpha.all)
car::leveneTest(integrity ~ kit, data = alpha.all)

bartlett.test(integrity ~ soil_type, data = alpha.all)
fligner.test(integrity ~ soil_type, data = alpha.all)
car::leveneTest(integrity ~ soil_type, data = alpha.all)

shapiro.test(comp.bray$bray)
qqnorm(comp.bray$bray)
qqline(y = comp.bray$bray)

bartlett.test(bray ~ kit, data = comp.bray)
fligner.test(bray ~ kit, data = comp.bray)
car::leveneTest(bray ~ kit, data = comp.bray)

bartlett.test(bray ~ soil_type, data = comp.bray)
fligner.test(bray ~ soil_type, data = comp.bray)
car::leveneTest(bray ~ soil_type, data = comp.bray)

# ANOVA yield
aov.yield <- lm(yield_ng ~ kit * soil_type, data = alpha.all)

aov.yield %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.yield))$kit
TukeyHSD(aov(aov.yield))$soil_type
TukeyHSD(aov(aov.yield))$`kit:soil_type`

data.yield <- alpha.all %>% filter(!is.na(yield_ng)) %>% pull(yield_ng)
residuals.yield <- resid(aov.yield)
df.yield <- cbind(data.yield, residuals.yield) %>% as.data.frame(.)

ggplot(aes(x = data.yield,
           y = residuals.yield),
       data = df.yield) +
  geom_point() +
  geom_hline(yintercept = 0)

aov.yield.ranks <- lm(yield_ng_rank ~ kit * soil_type, data = alpha.all)

aov.yield.ranks %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.yield.ranks))$kit
TukeyHSD(aov(aov.yield.ranks))$soil_type
TukeyHSD(aov(aov.yield.ranks))$`kit:soil_type`

data.yield.rank <- alpha.all %>% filter(!is.na(yield_ng_rank)) %>% pull(yield_ng_rank)
residuals.yield.rank <- resid(aov.yield.ranks)
df.yield.rank <- cbind(data.yield.rank, residuals.yield.rank) %>% as.data.frame(.)

ggplot(aes(x = data.yield.rank,
           y = residuals.yield.rank),
       data = df.yield.rank) +
  geom_point() +
  geom_hline(yintercept = 0)

# Non-parametric test for differences in purity
ratio1.PS_LT <- alpha.all %>% filter(kit == "PowerSoil LT") %>% pull(`260/280`)

ratio1.PS_HT <- alpha.all %>% filter(kit == "PowerSoil Pro HT") %>% pull(`260/280`)

ratio2.PS_LT <- alpha.all %>% filter(kit == "PowerSoil LT") %>% pull(`260/230`)

ratio2.PS_HT <- alpha.all %>% filter(kit == "PowerSoil Pro HT") %>% pull(`260/230`)

test.ratio1 <- wilcox.test(ratio1.PS_LT, ratio1.PS_HT, alternative = "less")
test.ratio1$p.value
# Add small jitter due to ties
d1 = ratio1.PS_LT-ratio1.PS_HT
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val
j = runif(15, -.01, .01)
dj1 = d1+j; wilcox.test(dj1)$p.val

test.ratio2 <- wilcox.test(ratio2.PS_LT, ratio2.PS_HT, alternative = "less")
test.ratio2$p.value
# Add small jitter due to ties
d2 = ratio2.PS_LT-ratio2.PS_HT
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val
j = runif(15, -.01, .01)
dj2 = d2+j; wilcox.test(dj2)$p.val

p.adj_ratio <- p.adjust(c(test.ratio1$p.value, test.ratio2$p.value), method = "bonferroni")
p.adj_ratio

# ANOVA integrity
aov.integrity <- lm(integrity ~ kit * soil_type, data = alpha.all)

aov.integrity %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.integrity))$kit
TukeyHSD(aov(aov.integrity))$soil_type
TukeyHSD(aov(aov.integrity))$`kit:soil_type`

data.integrity <- alpha.all %>% filter(!is.na(integrity)) %>% pull(integrity)
residuals.integrity <- resid(aov.integrity)
df.integrity <- cbind(data.integrity, residuals.integrity) %>% as.data.frame(.)

ggplot(aes(x = data.integrity,
           y = residuals.integrity),
       data = df.integrity) +
  geom_point() +
  geom_hline(yintercept = 0)

aov.integrity.ranks <- lm(integrity_rank ~ kit * soil_type, data = alpha.all)

aov.integrity.ranks %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.integrity.ranks))$kit
TukeyHSD(aov(aov.integrity.ranks))$soil_type
TukeyHSD(aov(aov.integrity.ranks))$`kit:soil_type`

plot(TukeyHSD(aov(aov.integrity.ranks))$kit, col="black", cex.axis=0.75)

data.integrity.rank <- alpha.all %>% filter(!is.na(integrity_rank)) %>% pull(integrity_rank)
residuals.integrity.rank <- resid(aov.integrity.ranks)
df.integrity.rank <- cbind(data.integrity.rank, residuals.integrity.rank) %>% as.data.frame(.)

ggplot(aes(x = data.integrity.rank,
           y = residuals.integrity.rank),
       data = df.integrity.rank) +
  geom_point() +
  geom_hline(yintercept = 0)

# Non-parametric test for differences in integrity
integrity.FS_LT <- alpha.all %>% filter(kit == "FastSpin LT") %>% pull(integrity)
integrity.PS_LT <- alpha.all %>% filter(kit == "PowerSoil LT") %>% pull(integrity)
integrity.FS_HT <- alpha.all %>% filter(kit == "FastSpin HT") %>% pull(integrity)
integrity.PS_HT <- alpha.all %>% filter(kit == "PowerSoil Pro HT") %>% pull(integrity)
integrity.ZM_HT <- alpha.all %>% filter(kit == "ZymoMagbead HT") %>% pull(integrity)

shapiro.test(integrity.FS_LT)
qqnorm(integrity.FS_LT)
qqline(integrity.FS_LT)

shapiro.test(integrity.PS_LT)
qqnorm(integrity.PS_LT)
qqline(integrity.PS_LT)

shapiro.test(integrity.FS_HT)
qqnorm(integrity.FS_HT)
qqline(integrity.FS_HT)

shapiro.test(integrity.PS_HT)
qqnorm(integrity.PS_HT)
qqline(integrity.PS_HT)

shapiro.test(integrity.ZM_HT)
qqnorm(integrity.ZM_HT)
qqline(integrity.ZM_HT)

median.integrity <- c(median(integrity.FS_LT), median(integrity.PS_LT), median(integrity.FS_HT), median(integrity.PS_HT), median(integrity.ZM_HT))
round(median.integrity/1000, 1)

mean.integrity <- c(mean(integrity.FS_LT), mean(integrity.PS_LT), mean(integrity.FS_HT), mean(integrity.PS_HT), mean(integrity.ZM_HT))
round(mean.integrity/1000, 1)

sd.integrity <- c(sd(integrity.FS_LT), sd(integrity.PS_LT), sd(integrity.FS_HT), sd(integrity.PS_HT), sd(integrity.ZM_HT))
round(sd.integrity/1000, 1)

test.integrity <- wilcox.test(integrity.PS_HT, integrity.PS_LT, alternative = "less")
test.integrity$p.value
test.ratio1$p.value
# Add small jitter due to ties
d3 = integrity.PS_HT-integrity.PS_LT
j = runif(15, -.01, .01)
dj3 = d3+j; wilcox.test(dj3)$p.val
j = runif(15, -.01, .01)
dj3 = d3+j; wilcox.test(dj3)$p.val
j = runif(15, -.01, .01)
dj3 = d3+j; wilcox.test(dj3)$p.val
j = runif(15, -.01, .01)
dj3 = d3+j; wilcox.test(dj3)$p.val
j = runif(15, -.01, .01)
dj3 = d3+j; wilcox.test(dj3)$p.val

# ANOVA shannon
aov.shannon <- lm(shannon ~ kit * soil_type, data = alpha.all)

aov.shannon %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.shannon))$kit

data.shannon <- alpha.all %>% filter(!is.na(shannon)) %>% pull(shannon)
residuals.shannon <- resid(aov.shannon)
df.shannon <- cbind(data.shannon, residuals.shannon) %>% as.data.frame(.)

ggplot(aes(x = data.shannon,
           y = residuals.shannon),
       data = df.shannon) +
  geom_point() +
  geom_hline(yintercept = 0)

aov.shannon.ranks <- lm(shannon_rank ~ kit * soil_type, data = alpha.all)

aov.shannon.ranks %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.shannon.ranks))$kit

data.shannon.rank <- alpha.all %>% filter(!is.na(shannon_rank)) %>% pull(shannon_rank)
residuals.shannon.rank <- resid(aov.shannon.ranks)
df.shannon.rank <- cbind(data.shannon.rank, residuals.shannon.rank) %>% as.data.frame(.)

ggplot(aes(x = data.shannon.rank,
           y = residuals.shannon.rank),
       data = df.shannon.rank) +
  geom_point() +
  geom_hline(yintercept = 0)

# ANOVA bray
aov.bray <- lm(bray ~ kit * soil_type, data = comp.bray)

aov.bray %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.bray))$kit

data.bray <- comp.bray %>% filter(!is.na(bray)) %>% pull(bray)
residuals.bray <- resid(aov.bray)
df.bray <- cbind(data.bray, residuals.bray) %>% as.data.frame(.)

ggplot(aes(x = data.bray,
           y = residuals.bray),
       data = df.bray) +
  geom_point() +
  geom_hline(yintercept = 0)

aov.bray.ranks <- lm(bray_rank ~ kit * soil_type, data = comp.bray)

aov.bray.ranks %>%
  anova(.) %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(R2 = Sum.Sq / sum(Sum.Sq) * 100) %>%
  mutate(across(R2, ~ as.character(round(., digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

TukeyHSD(aov(aov.bray.ranks))$kit

data.bray.rank <- comp.bray %>% filter(!is.na(bray_rank)) %>% pull(bray_rank)
residuals.bray.rank <- resid(aov.bray.ranks)
df.bray.rank <- cbind(data.bray.rank, residuals.bray.rank) %>% as.data.frame(.)

ggplot(aes(x = data.bray.rank,
           y = residuals.bray.rank),
       data = df.bray.rank) +
  geom_point() +
  geom_hline(yintercept = 0)
```


## Heatmap
```{r}
# Find all phyla in amplicon data
# heatmap.amplicon_phylum <- dat.types.sub10000.amp %>%
#   amp_heatmap(group_by = "seq_id", 
#               min_abundance = 0.001,
#               tax_show = 1000,
#               tax_aggregate = "Phylum", 
#               normalise = FALSE,
#               showRemainingTaxa = FALSE, 
#               plot_values = TRUE,
#               textmap = TRUE) %>%
#   rownames_to_column(var = "Phylum") %>%
#   mutate(across(Phylum, ~ str_replace_all(., "^.*ASV.*$", "Unclassified"))) %>%
#   group_by(Phylum) %>%
#   summarise(across(everything(), ~ sum(.))) %>%
#   ungroup() %>%
#   column_to_rownames(var = "Phylum")
# 
# phylum.text.amp <- heatmap.amplicon_phylum %>%
#   mutate(total = rowSums(.)) %>%
#   filter(!total == 0) %>%
#   select(total) %>%
#   arrange(desc(total)) %>%
#   row.names(.)
# 
# phylum.text.renamed <- gsub("SAR324_cladeMarine_group_B", "\\*SAR324", phylum.text.amp)

# Load new genus ampvis object
# ampvis.sub_phylum <- amp_load(otutable = heatmap.amplicon_phylum,
#                               metadata = metadata.amplicon)

# plot.heatmap_phylum <- ampvis.sub_phylum %>%
#   #amp_subset_samples(., normalise = TRUE) %>%
#   amp_heatmap(group_by = "kit", 
#               min_abundance = 0.01,
#               facet_by = "soil_type",
#               tax_aggregate = "OTU",
#               normalise = TRUE,
#               #tax_show = 25,
#               tax_show = phylum.text.amp[c(1:12, 14:26)], 
#               order_y_by = rev(phylum.text.amp[c(1:12, 14:26)]),
#               showRemainingTaxa = FALSE, 
#               plot_na = TRUE,
#               plot_values = TRUE,
#               plot_values_size = 3,
#               round = 2,
#               color_vector = c("white", colors[3]),
#               plot_legendbreaks = c(0.01, 0.1, 1, 10, 30),
#               plot_colorscale = "log10") +
#   scale_y_discrete(labels = rev(c(phylum.text.renamed[c(1:12, 14:26)]))) +
#   scale_x_discrete(labels = kit_levels) +
#   labs(x = "", 
#        y = "", 
#        tag = "A",
#        title = "Heatmap aggregated at the phylum level across DNA extraction kits") +
#   articletheme  
#   #theme(plot.margin = unit(c(0, 1, 0, 0), "cm"))
  
plot.heatmap_phylum <- dat.types.sub10000.amp %>%
  #amp_subset_taxa(tax_vector = c("Actinobacteriota", "Firmicutes"), normalise = TRUE) %>%
  amp_subset_samples(normalise = TRUE) %>%
  #filter_otus(0.1) %>%
  amp_heatmap(group_by = "kit", 
              min_abundance = 0.001,
              facet_by = "soil_type",
              tax_aggregate = "Genus",
              #tax_add = c("Phylum"),
              normalise = FALSE,
              tax_show = 25,
              #order_y_by = "cluster",
              # tax_show = phylum.text.amp[c(1:12, 14:26)],
              # order_y_by = rev(phylum.text.amp[c(1:12, 14:26)]),
              showRemainingTaxa = FALSE,
              plot_na = TRUE,
              plot_values = FALSE,
              plot_values_size = 2,
              round = 1,
              #color_vector = c(colors[3], "white", colors[5]),
              plot_legendbreaks = c(0.01, 0.5, 2, 4, 6, 8),
              plot_colorscale = "sqrt") +
  scale_x_discrete(labels = kit_levels) +
  guides(fill = guide_colourbar(ticks.colour = "black")) +
  #scale_y_discrete(labels = rev(c(phylum.text.renamed[c(1:12, 14:26)]))) +
  labs(x = "", 
       y = "", 
       tag = "A",
       title = "Heatmap aggregated at the genus level across DNA extraction kits") +
  articletheme  
  #theme(plot.margin = unit(c(0, 1, 0, 0), "cm"))

plot.heatmap_phylum

#leg.plot_heat <- cowplot::get_legend(plot.heatmap_phylum)

plot.heatmap_phylum.final <- plot.heatmap_phylum + theme(legend.position = "none")

plot.heatmap_phylum

#ggsave(here("Benchmarking", "figures", "heatmap_benchmarking.png"), plot = plot.heatmap_phylum, device = "png", height = 12, width = 19)
```

# PERMANOVA test for overall difference in observed communities based on kit
```{r statistical testing}
## Overall comparison
permanova.list <- list()
      
for (i in 1:length(soil)) {
  
  community_0.1 <- dat.types.sub10000.amp %>%
    amp_subset_samples(normalise = TRUE) %>%
    filter_otus(0.1) %>%
    amp_subset_samples(soil_type == soil[i])
  
  community_0.1 <- t(community_0.1[["abund"]])
  
  meta <- dat.types.sub10000.amp[["metadata"]] %>%
    filter(soil_type == soil[i])
  
  dis <- vegan::vegdist(community_0.1)
  
  groups.stratified <- meta %>%
    mutate(group = str_c(soil_type, kit, sep = " ")) %>%
    pull(group) %>%
    as.factor(.)
  
 permanova <- adonis2(community_0.1 ~ kit, type = "bray", data = meta, permutations = 99999) %>%
   as.data.frame() %>%
   rownames_to_column("Variable") %>%
   mutate(across(R2, ~ as.character(round(. * 100, digits = 1)))) %>%
   mutate(pval = ifelse(`Pr(>F)` < 0.001, "< 0.001", round(`Pr(>F)`, 2))) %>%
   select(Variable, R2, pval)
 
 permanova.list[[i]] <- permanova
 
 names(permanova.list) <- soil[i]
}
  
names(permanova.list) <- soil

perm.results <- as.numeric(c(permanova.list[[1]][1,2], permanova.list[[2]][1,2], permanova.list[[3]][1,2], permanova.list[[4]][1,2], permanova.list[[5]][1,2]))

shapiro.test(perm.results)
mean(perm.results)
sd(perm.results)

# Statified PERMANOVA
# Betadispersion - analysis of multivariate homogeneity of group dispersions (variances)
community_0.1 <- dat.types.sub10000.amp %>%
  amp_subset_samples(normalise = TRUE) %>%
  filter_otus(0.1)

community_0.1 <- t(community_0.1[["abund"]])

meta <- dat.types.sub10000.amp[["metadata"]]

dis <- vegan::vegdist(community_0.1)

groups.stratified <- meta %>%
  mutate(group = str_c(soil_type, kit, sep = " ")) %>%
  pull(group) %>%
  as.factor(.)

mod.stratified <- vegan::betadisper(dis, groups.stratified)
mod.stratified

anova(mod.stratified)

plot(mod.stratified, ellipse = TRUE, hull = FALSE, conf = 0.90) # 90% data ellipse

perm.strat <- with(meta, permute::how(nperm = 99999, blocks = soil_type))

set.seed(123)
permanova.stratified <- adonis2(community_0.1 ~ kit*soil_type, type = "bray", data = meta, permutations = 99999)

permanova.stratified %>%
  data.frame() %>%
  rownames_to_column("Variable") %>%
  mutate(across(R2, ~ as.character(round(. * 100, digits = 1)))) %>%
  mutate(pval = ifelse(Pr..F. < 0.001, "< 0.001", round(Pr..F., 2))) %>%
  select(Variable, R2, pval) %>%
  gridExtra::tableGrob(rows = NULL) %>%
  gridExtra::grid.arrange(.)

#write_csv(bray_distance, here("Benchmarking", "figures", "2023-03-04_bray-distance.csv"))
#write_csv(comp.bray, here("Benchmarking", "figures", "2023-03-04_bray-comparisons.csv"))
```

## Ordination combined
```{r}
plot.ordinate <- dat.types.sub10000.amp %>%
  amp_ordinate(filter_species = 0.1,
               transform = "hellinger",
               distmeasure = "bray",
               sample_shape_by = "kit",
               #sample_colorframe = "soil_type",
               species_nlabels = 8,
               species_label_taxonomy = "Phylum",
               species_plot = T) +
  scale_color_manual(values = colors[c(1,3,4,5,2)]) +
  scale_fill_manual(values = colors[c(1,3,4,5,2)]) +
  scale_shape_manual(values = c(22,21,15:17), labels = kit_levels) +
  labs(title = "All samples: PCA of microbial diversity",
       fill = "Soil type",
       shape = "Extraction kit",
       tag = "A",
       subtitle = paste0("**PERMANOVA:**", 
                         "<br>Extraction kit: Variance explained = ", "2.8 %,", " p-value < 0.001",
                         "<br>Soil type: Variance explained = ", "90.6 %,", " p-value < 0.001",
                         "<br>Interaction: Variance explained = ", "4.5 %,", " p-value < 0.001")) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = soil_type), alpha = 0.5, color = "black") +
  coord_cartesian(xlim = c(-0.5, 0.5), ylim = c(-0.6, 0.4)) +
  articlethemex0 + 
  theme(plot.subtitle = ggtext::element_markdown())

plot.ordinate

#The plot with a zoom
plot.insert <- #plot.ordinate +
  dat.types.sub10000.amp %>%
  amp_ordinate(filter_species = 0.1,
               transform = "hellinger",
               distmeasure = "bray",
               sample_shape_by = "kit",
               #sample_colorframe = "soil_type",
               species_nlabels = 8,
               species_label_taxonomy = "Phylum",
               species_plot = T) +
  scale_color_manual(values = colors[c(1,3,4,5,2)]) +
  scale_fill_manual(values = colors[c(1,3,4,5,2)]) +
  scale_shape_manual(values = c(22,21,15:17), labels = kit_levels) +
    ggforce::geom_mark_ellipse(aes(fill = NULL, group = soil_type), alpha = 0.5, color = NA) +
  articlethemex0 + 
  coord_cartesian(ylim = c(-0.545, -0.56), xlim = c(0.185, 0.21)) +
  theme(plot.title = element_blank(),
        plot.subtitle = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        panel.background = element_rect(fill = alpha(colors[3], 0.5)),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        #panel.background = element_blank(),
        legend.position = "none",
        axis.ticks = element_blank(),
        plot.margin = unit(c(0, 0, 0, 0), "cm"),
        plot.tag = element_blank())

plot.insert$layers[[3]] <- NULL

plot.insert 

plot.ordinate.insert <- plot.ordinate + 
  #geom_segment(aes(x = 0.19, y = 0.57, xend = -0.3, yend = 0.6), color = "black", linetype = "dashed") +
  geom_segment(aes(x = 0.185, y = -0.56, xend = -0.1, yend = -0.62), color = "black", linetype = "dashed") +
  #geom_segment(aes(x = 0.19, y = 0.54, xend = -0.3, yend = 0.25), color = "black", linetype = "dashed") +
  geom_segment(aes(x = 0.185, y = -0.545, xend = -0.1, yend = -0.225), color = "black", linetype = "dashed") +
  annotation_custom(ggplotGrob(plot.insert), xmin = -0.5, xmax = -0.1, ymin = -0.225, ymax = -0.625) +
  geom_rect(aes(xmin = 0.185, xmax = 0.21, ymin = -0.545, ymax = -0.56), color='black', alpha = 0) #+
  #geom_rect(aes(xmin = -0.275, xmax = 0.075, ymin = 0.25, ymax = 0.6), color = "black", linetype = "dashed", alpha = 0)

plot.ordinate.insert

leg.plot.ord_shape <- cowplot::get_legend(plot.ordinate.insert + guides(fill = "none"))

leg.plot.ord_fill <- cowplot::get_legend(plot.ordinate.insert + guides(shape = "none"))

plot.ordinate.insert_final <- plot.ordinate.insert + labs(tag = "B") + theme(legend.position = "none")

#ggsave(here("Benchmarking", "figures", "PCA_ordination_insert.png"), plot = plot.ordinate.insert, device = "png", height = 12, width = 19)
```



## Ordinations - faceted by soil types
```{r ordination}
plot_ord1 <- dat.types.sub10000.amp %>%
  amp_subset_samples(soil_type == soil[1]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "kit",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = kit), alpha = 0.5, color = "black") +
  scale_fill_manual(values = c(rep(colors[1], 5))) +
  scale_shape_manual(values = c(22,21,15:17), labels = kit_levels) +
  labs(title = paste0(soil[1], ": ", "PCA of microbial diversity"),
       tag = "B",
       subtitle = paste0("**PERMANOVA:**", "<br>Extraction kit: Variance explained = ", permanova.list[[1]][1,2], " %,", " p-value ", permanova.list[[1]][1,3], "<br>", "<br>")) +
  coord_cartesian(xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5)) +
  articlethemex0 +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())
  
plot_ord2 <- dat.types.sub10000.amp %>%
  amp_subset_samples(soil_type == soil[2]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "kit",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = kit), alpha = 0.5, color = "black") +
  scale_fill_manual(values = c(rep(colors[3], 5))) +
  scale_shape_manual(values = c(22,21,15:17), labels = kit_levels) +
  labs(title = paste0(soil[2], ": ", "PCA of microbial diversity"),
       tag = "C",
       subtitle = paste0("**PERMANOVA:**", "<br>Extraction kit: Variance explained = ", permanova.list[[2]][1,2], " %,", " p-value ", permanova.list[[2]][1,3], "<br>", "<br>")) +
  coord_cartesian(xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5)) +
  articlethemex0 +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())

plot_ord3 <- dat.types.sub10000.amp %>%
  amp_subset_samples(soil_type == soil[3]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "kit",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = kit), alpha = 0.5, color = "black") +
  scale_fill_manual(values = c(rep(colors[4], 5))) +
  scale_shape_manual(values = c(22,21,15:17), labels = kit_levels) +
  labs(title = paste0(soil[3], ": ", "PCA of microbial diversity"),
       tag = "D",
       subtitle = paste0("**PERMANOVA:**", "<br>Extraction kit: Variance explained = ", permanova.list[[3]][1,2], " %,", " p-value ", permanova.list[[3]][1,3], "<br>", "<br>")) +
  articlethemex0 +
  coord_cartesian(xlim = c(-0.75, 0.25), ylim = c(-0.5, 0.5)) +
  articlethemex0 + 
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())

plot_ord4 <- dat.types.sub10000.amp %>%
  amp_subset_samples(soil_type == soil[4]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "kit",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = kit), alpha = 0.5, color = "black") +
  scale_fill_manual(values = c(rep(colors[5], 5))) +
  scale_shape_manual(values = c(22,21,15:17), labels = kit_levels) +
  labs(title = paste0(soil[4], ": ", "PCA of microbial diversity"),
       tag = "E",
       subtitle = paste0("**PERMANOVA:**", "<br>Extraction kit: Variance explained = ", permanova.list[[4]][1,2], " %,", " p-value ", permanova.list[[4]][1,3], "<br>", "<br>")) +
  coord_cartesian(xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5)) +
  articlethemex0 +
  articlethemex0 +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())

plot_ord5 <- dat.types.sub10000.amp %>%
  amp_subset_samples(soil_type == soil[5]) %>%
  amp_ordinate(
    filter_species = 0.1,
    sample_shape_by = "kit",
    type = "PCA",
    transform = "hellinger",
    distmeasure = "bray",
    species_nlabels = 5,
    species_label_taxonomy = "Phylum",
    species_plot = T) +
  ggforce::geom_mark_ellipse(aes(fill = soil_type, group = kit), alpha = 0.5, color = "black") +
  scale_fill_manual(values = c(rep(colors[2], 5))) +
  scale_shape_manual(values = c(22,21,15:17), labels = kit_levels) +
  labs(title = paste0(soil[5], ": ", "PCA of microbial diversity"),
       tag = "F",
       subtitle = paste0("**PERMANOVA:**", "<br>Extraction kit: Variance explained = ", permanova.list[[5]][1,2], " %,", " p-value ", permanova.list[[5]][1,3], "<br>", "<br>")) +
  coord_cartesian(xlim = c(-0.5, 0.5), ylim = c(-0.5, 0.5)) +
  articlethemex0 +
  articlethemex0 +
  theme(legend.position = "none",
        plot.subtitle = ggtext::element_markdown())

#leg.plot_ord <- cowplot::get_legend(plot_ord)

plot_ord_leg <- ggarrange(leg.plot.ord_fill, leg.plot.ord_shape, nrow = 1, align = "hv")

plot.ord <- ggarrange(plot.ordinate, plot_ord1, plot_ord2, plot_ord3, plot_ord4, plot_ord5, common.legend = TRUE, ncol = 3, nrow = 2, legend = "top")

plot.ord

plot_ord_final <- plot_ord2 +
  labs(tag = "C") +
  theme(legend.position = "none")

ggsave(here("Benchmarking", "figures", "pre_S2Fig.png"), plot = plot.ord, device = "png", height = 12, width = 19)
ggsave(here("Benchmarking", "figures", "pre_S2Fig.tif"), plot = plot.ord, device = "tiff", height = 12, width = 19, dpi = 600)
```

## Final plot
```{r final plot}
# Create combined plot
plot <- ggarrange(plot.ordinate.insert_final, plot_ord_final, ncol = 1, nrow = 2)

plot

leg.heat <- cowplot::get_legend(plot.heatmap_phylum)
leg.ord <- cowplot::get_legend(plot.ordinate.insert)

leg.comb <- cowplot::plot_grid(leg.heat, leg.ord, ncol = 1, align = "v", nrow = 2, rel_heights = c(1, 1))

leg.comb

gg_arranged <- ggarrange(plot.heatmap_phylum.final, plot, leg.comb, ncol = 3, nrow = 1, widths = c(2, 1, 0.5))

gg_arranged

ggsave(here("Benchmarking", "figures", "pre_Fig2.png"), plot = gg_arranged, device = "png", height = 12, width = 19)
ggsave(here("Benchmarking", "figures", "pre_Fig2.tif"), plot = gg_arranged, device = "tiff", height = 12, width = 19, dpi = 600)
```




